How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).

### 2013-08-08 documents/turtle_paper.Rnw
results updated completely. some intro work. grow turtle section.

options(digits = 4)
@

@@

% cryptic diversity
%   most species are still deliminated solely based on morphology
Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}. Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis. This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citet{Hunt2008,Eldredge1972,Gould1977a}. In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups. Additionally, this would allow for better classifying extinct taxa.

Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted for case of purely morphological data. The majority of this effort has focused on identifying differences between already identified taxa \citep{Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
% all the work done on species delimitation strictly from molecular data

@@

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004}. These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides in interpretation and presentation of information. Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications \citep{MacLeod2007}. Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.

@@

% e. marmorata
%   natural history
%     conservation importance (turtles in general, e. marmorata in specific)
In this study, we investiage the subspecific classification of the western pond turtle, \textit{Emys marmorata}. \textit{E. marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
%   morphological hypothesis of subspecies
%   molecular hypothesis
Traditionally, \textit{E. marmorata} was classified into three groups: the northern \textit{E. marmorata marmorata}, the southern \textit{E. marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945,Holland1992}. \textit{E. marmorata marmorata} is differentiated from \textit{E. marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings. It should be noted that the triangular inguinal plates can sometimes be present in \textit{E. marmorata palida} though they are considerably smaller.

Previous work on morphological differentiation between different populations of \textit{E. marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}. 
%
\citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E. marmorata} from different populations over the entire range of the species. This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E. marmorata}. Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range. Additionally, sexual dimorphism present in \textit{E. marmorata} was only assesed in the context of size measures such as length of carapace and mass. In general, \citet{Holland1992} and others \citep{Lubcke2007,Germano2008} found that males of \textit{E. marmorata} are larger than females in terms of size. However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a} and \citet{Dryden1998a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.

@@

\section{Results}
\subsection{Geometric morphometrics}
The results of the PCA of the total dataset of \textit{E. marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig. \ref{fig:pca}). The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points with no structure. Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig. \ref{fig:pca}). However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.

The first two PCs appear to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig. \ref{fig:pc_var}). Landmark 10 (Fig. \ref{fig:plastra}), which appears to be the most variable along this axis (Fig. \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace. Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
Additional variation along the second PC appears to be characterized by variation in anterior and posterior landmarks along the medial edge of the plastron.

@@

\subsubsection{Supervised learning}
%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig. \ref{fig:roc}) methods include many, if not most, of the possible features.
<<>>=
daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
                      function(x) x[2]))
@
% multinomial logistic regression model
The AICc best multinomial logistic regression model for three of the four classification schemes include the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}). The second molecularly based classification hypothesis includes all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs predictors (Tables \ref{tab:mod_sel_4}).
The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}). The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{paste0(sum(turtle.fit$percent[1:9]))}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{paste0(sum(turtle.fit$perfect[1:10]))}\% of the variation.

While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model. While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the optimal model for three of the classification schemes having the same number of predictors and the fact that the \(\Delta\)AICc best model in these cases is the simpler mode as reasons to use only the AICc best model for all cases. 
Additionally, by using a single model for all classification schemes, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).

@@

The selected number of features in the final random forest model for each classification scheme varied as much as the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig. \ref{fig:roc}).

@@

In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1},\ref{tab:mod_sel_2},\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig. \ref{fig:roc}).

The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that the second molecular classification hypotheses was the best overall classification scheme (Fig. \ref{fig:gen_res}). For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}). Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.

@@

The first two most important features describe different aspects of variation (Fig. \ref{fig:imp_var}). The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron. Additionally, the relative position of the landmarks along the midline. This PC represents \Sexpr{turtle.fit$percent[3]}\% of total variation. The second and second most important PC is described above and principally described variation in landmarks along the medial and anterior edge of the plastron. This PC represents \Sexpr{turtle.fit$percent[2]}\% of total variation. The major variations along these axes correspond well to the differences between the mean shape of each class (Fig. \ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together.

The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute differently to classification and that given multiple features the odds of determining the correct classification increase (Fig. \ref{fig:rel_risk}). The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group. This is observable from the class histograms of PC 2 (Fig. \ref{fig:imp_var}). Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig. \ref{fig:rel_risk}) and the different class histograms of the PCs (Fig. \ref{fig:imp_var}).

@@

% methodological concerns
%   compromise in the supervised learning models
%     fair because as much variation as ``necessary'' is used
The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable. While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response. Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other. However, this compromise is not without its advantages. Because both AICc and AUC values improved rapidly with increased model complexity (Fig. \ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was. Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig. \ref{fig:rel_risk}). The results of the relative risk values also shows that while a variable can be considered important for the random forest models (Fig. \ref{fig:imp_pc}, \ref{fig:imp_var}) they each might not be considered useful on their own for all classes (Fig. \ref{fig:rel_risk}).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/imp_var}
  \caption{Landmark variation along the two most important features (PCs) based on the final random forest model. The first row corresponds to the third PC and the second corresponds to eighth PC. Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}
  \label{fig:imp_var}
\end{figure}

>>>>

options(digits = 2)

#knit_hooks$set(inline = function(x) {
#               if (is.numeric(x)) round(x, 4)})

@

@@

% cryptic diversity
%   most species are still deliminated solely based on morphology
Cryptic diversity is the phenomenon that not all taxa can be recognized based on morphology and can only be delimited using molecular information \citep{Stuart2006,Pfenninger2007,Funk2012,Clare2011}. Conceringly, most extant taxa, and nearly all extinct taxa, are delimited based solely via morphological analysis. This phenomenon is of great concern when studying variation and diversity dynamics over long periods of time, where apparent morphological stasis may not actually reflect true diversity \citep{Hunt2008,Eldredge1972,Gould1977a}. In the case of endangered or conserved taxa, morphometric approaches for classifying and identifying taxa and populations of importance would greatly improve the ability to maintain these high risk groups. Additionally, this would allow for better classifying extinct taxa.

Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} while comparatively little has been devoted to introducing new methodology for case of purely morphological data \citep{Mitteroecker2011,Zelditch2004}. The majority of this effort has focused on identifying differences between already identified taxa \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}.
% all the work done on species delimitation strictly from molecular data

@@

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009}. These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides in interpretation and presentation of information. These previous studies, however, do not compare which amongst a set of candidate classification hypotheses is better. Also, there is no generalization of the classification model and training data accuracy is almost exclusively used as the metric off strength of classification.

Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. While machine learning methods such as neural networks have been applied to studying shape variation \citep{MacLeod2007}, they have been primarily applied in the context of automated taxon identification and not in terms of group classification and strength of classification. Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.

@@

% e. marmorata
%   natural history
%     conservation importance (turtles in general, e. marmorata in specific)
In this study, we investiage the subspecific classification of the western pond turtle, \textit{Emys marmorata}. \textit{E. marmorata} is distributed from northern Washington State, USA to Baja California, Mexico.
%   morphological hypothesis of subspecies
%   molecular hypothesis
Traditionally, \textit{E. marmorata} was classified into three groups: the northern \textit{E. marmorata marmorata}, the southern \textit{E. marmorata palida}, and a central Californian intergrade zone \citep{Seeliger1945}. \textit{E. marmorata marmorata} is differentiated from \textit{E. marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings. It should be noted that the triangular inguinal plates can sometimes be present in \textit{E. marmorata palida} though they are considerably smaller.

Previous work on morphological differentiation in \textit{E. marmorata} focused on comparisons between different populations within the rage CITATIONS. Most of these studies were not over the entire range and used only linear measures of morphology instead of multivariate shape CITATIONS.
%
\citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E. marmorata} from different populations over the entire range of the species. This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E. marmorata}. Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range. Additionally, sexual dimorphism present in \textit{E. marmorata} was only assesed in the context of size measures such as length of carapace and mass. In general, \citet{Holland1992} and others \citep{Lubcke2007,Germano2008} found that males of \textit{E. marmorata} are larger than females in terms of size. However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a} and \citet{Dryden1998a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.

@@

\section{Results}
\subsection{Geometric morphometrics}
The results of the PCA of the total dataset of \textit{E. marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig. \ref{fig:pca}). The first three PCs, which represent \Sexpr{sum(turtle.fit$percent[1:3])}\% of the total variation, are a cloud of points with no structure. Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig. \ref{fig:pca}). However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean shape \citep{Zelditch2004} and this observation should be considered cursory at best.

The first two PCs appear to describe principally varaition in the lateral margin of te palstra, from a pointed medial edge to a more rounded and blunt edge (Fig. \ref{fig:pc_var}). Landmark 10 (Fig. \ref{fig:plastra}), which appears to be the most variable along these axes (Fig. \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace. Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
Lateral landmark variation along the first PC seemed concentrated in the posterior portion of the plastra with additional variance observed in midline landmarks (Fig. \ref{fig:pc_var}). This variance in midline landmakrs was most likely caused by the fact that plastral scutes frequently do not line up perfectly. Along PC 2, lateral variation appeared to be concetrated in the anterior portion o the plastra (Fig. \ref{fig:pc_var}).

@@

\subsubsection{Supervised learning}
%For all of the classification schemes, the optimal models for both the multinomial logistic regression (Table SUPPLEMENT) and random forest (Fig. \ref{fig:roc}) methods include many, if not most, of the possible features.
<<>>=
daic <- unlist(lapply(lapply(tm.a.analysis$sel, function(x) x$delta), 
                      function(x) x[2]))
@
% multinomial logistic regression model
The AICc best multinomial logistic regression model for three of the four classification schemes had the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs as features (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, and \ref{tab:mod_sel_3}). The second molecularly based classification hypothesis included all \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} possible PCs as predictors (Tables \ref{tab:mod_sel_4}).
The \(\Delta\)AICc values between the optimal and second best model range from \Sexpr{min(daic)} for the first morphological based classification hypothesis to \Sexpr{max(daic)} for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}). The first \Sexpr{length(tm.a.analysis$best$spinks$coefnames) - 1} PCs describe \Sexpr{sum(turtle.fit$percent[1:9])}\% of total variation in plastral shape, while the first \Sexpr{length(tm.a.analysis$best$spinks$coefnames)} PCs describe \Sexpr{sum(turtle.fit$percent[1:10])}\% of the variation.

While the \(\Delta\)AICc value between the optimal and second best model for the first morphological and first molecular based classification hypothesis was within the range to be considered equally optimal \citep{Burnham2002a}, for this analysis we chose to use only the AICc best model. While AICc values can not be compared between models with different responses \citep{Burnham2002a}, we interpret the fact that the \(\Delta\)AICc best model in these cases is the simpler model and that the optimal model for three of the classification schemes having the same number of predictors as reasons to use only the AICc best model for all cases. 
Additionally, by using a single model for each of the classification hypotheses, this limits the number of comparisons between the bootstrap resampled distributions of the AUC values for the testing dataset (see below).

@@

The selected number of features in the final random forest model for each classification scheme was very simplar to the model selection results from the multinomial logistic regression models, ranging from \Sexpr{min(oprf)} for the second morphological based classification hypothesis and both molecular based classification hypotheses to \Sexpr{max(oprf)} for the first morphological based classification hypothesis (Fig. \ref{fig:roc}).

@@

In the case of all models, there is a substantial increase in model performance as measured by AICc for the multinomial logistic models (Tables \ref{tab:mod_sel_1}, \ref{tab:mod_sel_2}, \ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}) or in AUC for the random forest models and illustrated for the multinomial logistic regression models as the number of features increases (Fig. \ref{fig:roc}).

The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that a molecular classification hypotheses was the best overall classification scheme (Fig. \ref{fig:gen_res}). Remarkably, the best classification hypothesis was the second molecular classification hypothesis based on both the multinomial logistic regression and random forest models. For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}).

@@

The first two most important features describe different aspects of variation (Fig. \ref{fig:imp_var}). The third and most important PC describes variation roundedness of the medial portion of the plastron, both the anterior and posterior portions of the plastron. Additionally, the relative position of the landmarks along the midline varies greatly along PC3 (Fig. \ref{fig:imp_var}). This PC represents \Sexpr{turtle.fit$percent[3]}\% of total variation. The second and second most important PC is described above and principally described variation in landmarks along the lateral and anterior margin of the plastron. This PC represents \Sexpr{turtle.fit$percent[2]}\% of total variation. The major variations along these axes correspond well to the differences between the mean shape of each class (Fig. \ref{fig:mean_shape}) where major class differences seem based on the relative ballooning or shrinking of the anterior and posterior portions of the plastron together along with differential ``pinching'' of the midline landmarks.

The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that individual axes contribute to classification differently and that given multiple features the odds of determining the correct classification increase (Fig. \ref{fig:rel_risk}). The first most important axis contributes strongly to classifying both the western and southern groups while changes along the second most important axes contribute very little to increasing the odds of classification for all but the eastern group. This is observable from the class histograms of PC 3 and 2 (Fig. \ref{fig:imp_var}). Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig. \ref{fig:rel_risk}) and the different class histograms of the PCs (Fig. \ref{fig:imp_var}).

@@

% methodological concerns
%   compromise in the supervised learning models
%     fair because as much variation as ``necessary'' is used
The approaches presented here for supervised learning analysis of the landmark variation represent a compromise between explicitly modeling all shape variation and preventing models from being overfit and ungeneralizable. While all aspects of shape may be evolving simultaneously, and not along individual PCs, including all shape variation in each model might increase model complexity beyond a reasonable level for the sample size and possibly the necessary complexity to accurately model the response. Additionally, because only individual PCs are used as features in the models, this does not accurately represent shape evolution and how exactly different classes might be evolving in relation to each other. However, this compromise is not without its advantages. Because both AICc and AUC values improved rapidly with increased model complexity (Fig. \ref{fig:roc}), this helped demonstrate how fine scale the actual variation between classes was. The variable importance information from the random forest models was extremely useful for understanding what aspects shape variance contributed most to differentiating the classes and in what order as opposed purely in the order of largest variance (Fig. \ref{fig:imp_pc} and \ref{fig:imp_var}). Additionally, the relative risk values from the mulitinomial logistic regression models demonstrate that a single PC is probably not sufficient for estimating the class of an observation, but that given a set of PCs this classification would be more accurate (Fig. \ref{fig:rel_risk}).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/imp_var}
  \caption{Landmark variation along the two most important features (PCs) based on the final random forest model. The first row corresponds to the third PC and the second corresponds to the second PC. Landmark configurations are minimum observed on that PC, mean shape, and maximum observed on that PC.}
  \label{fig:imp_var}
\end{figure}
